Cluster Difference Imaging Photometric Survey. III. The Coeval Halo and Core of NGC\,2516

### 2021-01-27 cluster_paper/ms.tex
more words

\defcitealias{bouma_wasp4b_2019}{B19}

@@

\keywords{
	stellar ages (1581),
	young star clusters (1833),
  open clusters (1160),
  stellar rotation (1629)
}

@@

How does the cluster evolve?  The answer depends
on the total stellar mass.
% What exactly does this mean?
The smallest 90\% of star clusters disperse during the ``embedded
phase'' (CITE).  This is driven mostly by {\it process A} (CITE).  The
more massive clusters make it maybe 100 Myr. Maybe a bit longer.
Their evaporation is thought to be driven by collisions with molecular
clouds (CITE, Spitzer 1958), and the galactic tide (CITE), and {\it
process A}, and {\it process B}.  Of course, the expected evaporation
time should depend on factors including the mass of the cluster
itself, and how many high-mass (O and B) stars form, since their winds
[and maybe supernovae] clear out most of the cloud, and the initial
density of the gas cloud to begin with.

@@

Identification of the stars that disperse into the galactic field is
an important task for understanding the conditions under which stars
and star clusters form, and for understanding how they subsequently
evolve.  For instance, how does the process of radial migration across
the galactic disk affect cluster dispersal?  Did the Sun form in an
open cluster? If so, how massive was its cluster, and is there any
hope at identifying the stars that formed near the Sun?
Qualitatively, the dispersal of open clusters also provides perhaps
the best ``test case'' for the concept of ``chemical tagging'', also
referred to as ``galactic archaeology'' (CITE).

Outside of the issue of star formation, identifying the remnant halos
of open clusters is important for a separate project: that of
discovering young transiting planets.  Young transiting planets are
hard to find because young stars are rare (CITE), and often reside in
crowded regions of the sky (CITE).  If the halos of nearby star
clusters can be reliably identified, this could expand the census of
nearby sub-Gyr stars by a factor of 2 or even 3 (CITE Meingast).  A
fortuitous benefit of searching for planets in cluster environments is
that issues with stellar crowding are also alleviated.

Recent clustering studies using Gaia have begun to report the
identification of structures that could correspond to low-density
halos of stars that may have evaporated from open clusters (e.g.,
Kounkel+19, Kounkel+20, Meingast+21).  However, different clustering
methods on the Gaia data tend to give different results (CITE: Hunt \&
Reffert 2020).  Employing say Gaussian Mixture Modelling, or any
analogous method that requires ``clusters'' to be ellipses in
spatial/velocity phase-space unsurpsingly yields open clusters that
are roughly elliptical (CITE: Cantat-Gaudin+18). Unsupervised
clustering methods such as HDBScan (e.g., Kounkel+19) have been found
to yield additional structures, particularly in lower density regions
such as the Psc-Eri stream (CITE: Meingast+18, Curtis+19, Newton+21),
but also more generally around many open clusters (Kounkel+19).
Unsupervised approaches that incorporate physical constraints (e.g.,
imposing a maximum velocity dispersion on putative members) yield
similar results, potentially with higher purity (Meingast+21).

We've recently been making TESS light curves of age-dated stars across
the sky, as part of a Cluster Difference Imaging Photometry Survey
(CDIPS, CITE Bouma+19).  Our analysis of Cycle 1 (Sectors 1-13)
yielded light curves of \numsouthernuniqlcs candidate cluster members
in the Southern Ecliptic hemisphere, available on MAST\footnote{\url{https://archive.stsci.edu/hlsp/cdips}}.
% TODO: make this number actually quantitative...
Based on rotation periods, $\approx$25\% appear to be bonafide cluster
members.

As part of a broader project of identifying a large and clean sample
of young stars for a transit search, we focus in his
paper on a rather modest question: in just a single rich
southern open cluster, is the cluster halo coeval with the core?  The
cluster we chose for this analysis was NGC\,2516, since it was young
($\sim$112\,Myr) and close ($d=400$\,pc) enough to facilitate rotation
measurements using TESS, and some degree of spectroscopic analysis.
We want to know: is the halo real? To what extent can we use Gaia
alone to reliably identify age-dated needles in the haystack of boring
field stars?  And more generally, what are the implications for the
evolution of open clusters if they do have halos?

Section~\ref{sec:gaia} presents the astrometric and photometric data
from Gaia, and clarifies our usage of the terms ``core'' and ``halo''.
Section~\ref{sec:agedate} age-dates the halo of NGC\,2516,
based on Gaia photometry (Section~\ref{subsec:hr}),
TESS gyrochronology (Section~\ref{subsec:tess}), and lithium depletion
(Section~\ref{subsec:lithium}).  In Section~\ref{sec:discussion} we
discuss the implications of this analysis for NGC\,2516 specifically
and stellar spin-down and open cluster evolution generally.
Section~\ref{sec:conclusion} presents our conclusions.

@@

Figure~\ref{fig:gaia6d} shows the problem we would like to address.
In this figure, the ``core'' comprises \cn members that
\citet{cantatgaudin_gaia_2018} reported to have ``membership
probability'' exceeding 10\%, based on the Gaia DR2 astrometric data.
The exact meaning of this probability, and details concerning their
clustering algorithm, are discussed in Appendix~\ref{app:clustering}.
The ``halo'' comprises \cn members that
\citet{kounkel_untangling_2019} reported as members, also based on the
Gaia DR2 data.  That study did not report continuous membership
probabilities, instead opting for the binary ``member'' or ``not''.
These two different clustering methods yielded wildly different
results.  What is the true structure of \cn?
Are the core and halo truly coeval?

@@

The first check on whether this membership assignment is plausible was
already performed by \citet{kounkel_untangling_2019} and more recently
by \citet{meingast_2021}.  That check is to see whether the HR
diagrams of these cluster components that were selected based on
positions and velocities photometrically support the claim that they
are coeval.

@@

%
% TODO: the following two paragraphs, when quoted, are ENTIRELY FROM
% the relevant papers. they should probably be nixed in favor of
% concision.
%
\citet{cantatgaudin_gaia_2018} applied a procedure that ultimately
yielded what we call ``the core''.  Their procedure was to first query
a Gaia DR2 cone around the previously reported RA and dec of the
cluster, and within $\pm 0.5$\,mas of its previously reported
distance. 
% TODO: which was it?
The outer radius of their cone was either r2 from MWSC (CITE
Kharchenko 2013), or twice the ``cluster radius'' listed by Dias.  No
proper motion cut was applied.  They then applied an unsupervised
classification scheme called UPMASK to $G<18$\,mag stars within this
cone (CITE).  The steps of UPMASK are first to perform a k-means
clustering in the ``astrometric space'' ($\mu_{\alpha'}, \mu_\delta,
\pi$).  Then, a ``veto'' step is applied to assess whether the groups
of stars output from the k-means clustering are or are not more
concentrated than a random distribution.  This is implemented by
``comparing the total branch length of the minimum spanning tree
connecting the stars with the expected branch length for a random
uniform distribution covering the investigated field of view''.  ``To
turn this yes/no flag to a membership probability,
\citet{cantatgaudin_gaia_2018} then redraw new values of
($\mu_{\alpha'}, \mu_\delta, \pi$) for each source based on the listed
value, uncertainty, and covariance.  After a certain number of
redrawings, the final probability is the frequency with which a given
star passes the veto".  In the case of \cn, this yielded a reported
``\texttt{r50}'' within which half of the cluster members were found
to be within 0.496$^\circ$.

\citet{kounkel_untangling_2019} applied a different unsupervised
clustering method to the 5-dimensional Gaia DR2 data (omitting radial
velocities, due to their sparsity).  Their selection function (see
their Section~2) yielded $\approx 2\times 10^7$ stars, mostly within
$\approx 1$\,kpc and typically with $G<18$\,mag.  Their clustering
algorithm was the ``hierarchical density-based spatial clustering of
applications with noise'' (HDBSCAN, CITE McInnes17).  The classical
DBSCAN algorithm ``identifies clusters as overdensities in a
multi-dimensional space in which the number of sources exceeds the
required minimum number of points within a neighborhood of a
particular linking length $\epsilon$.  HDBSCAN does not depend on
$\epsilon$; instead it condenses the minimum spanning tree by pruning
off the nodes that do not meet the minimum number of sources in a
cluster and reanalyzing the nodes that do. Depending on the chosen
algorithm, it would then either find the most persistent structure
(through the excess of mass method), or return clusters as the leaves
of the tree (which results in somewhat more homogeneous clusters). In
both cases it is more effective at finding structures of varying
densities in a given data set than DBSCAN.''
`` The two main parameters that control HDBSCAN are the number of
sources in a cluster and the number of samples. The former is the
parameter that rejects groupings that are too small; the latter sets
the threshold of how conservative the algorithm is in its
considerations of the background noise (even if the resulting noisy
groupings do meet the minimum cluster size).  By default, the sample
size is set to the same value as the cluster size, but it is possible
to adjust them separately.''

>>>>

\defcitealias{bouma_wasp4b_2019}{B19}
\defcitealias{cantatgaudin_gaia_2018}{CG18}
\defcitealias{kounkel_untangling_2019}{KC19}

@@

\keywords{
	stellar ages (1581),
  stellar associations (1582),
  open clusters (1160),
  %stellar dynamics (1596),
  stellar rotation (1629)
}

@@

How does the cluster evolve?  The answer depends on the total stellar
mass.  The smallest 90\% of star clusters disperse during the ``embedded
phase'' (CITE Lada \& Lada 03, Krumholz 19 ARAA).  This is driven mostly
by {\it process A} (CITE).  The more massive clusters make it maybe 100
Myr. Maybe a bit longer.  Their evaporation is thought to be driven by
collisions with molecular clouds (CITE, Spitzer 1958, maybe Sec 7.2 of
Ryden), and the galactic tide (CITE), and {\it process A}, and {\it
process B}.  Of course, the expected evaporation time should depend on
factors including the mass of the cluster itself, and how many high-mass
(O and B) stars form, since their winds [and maybe supernovae] clear out
most of the cloud, and the initial density of the gas cloud to begin
with.

@@

Identification of the stars that disperse into the galactic field is an
important task for understanding the conditions under which stars and
star clusters form, and for understanding how they subsequently evolve.
For instance, how does the process of radial migration across the
galactic disk affect cluster dispersal?  Did the Sun form in an open
cluster? If so, how massive was its cluster, and is there any hope at
identifying the stars that formed near the Sun?  Qualitatively, the
dispersal of open clusters also provides perhaps the best test case for
the concept of ``chemical tagging'', aka.\ ``galactic archaeology''
(CITE).  If you can't chemically tag stars that are kinematically
associated, why would you be able to do so once they've dispersed
throughout the galactic disk?

Outside of the issue of star formation, identifying the remnant halos of
open clusters is important for a separate project: that of discovering
young transiting planets.  Young transiting planets are hard to find
because young stars are rare (CITE), and often reside in crowded regions
of the sky (CITE).  If the halos of nearby star clusters can be reliably
identified, this could expand the census of nearby sub-Gyr stars by a
factor of 2 or 3 (CITE).  A fortuitous benefit of searching for planets
in cluster environments is that issues with stellar crowding are also
alleviated.

Recent clustering studies using Gaia have begun to report the
identification of structures that could correspond to low-density halos
of stars that may have evaporated from open clusters (e.g., Kounkel+19,
Kounkel+20, Meingast+21).  However, different clustering methods on the
Gaia data tend to give different results (CITE: Hunt \& Reffert 2020).
Employing say Gaussian Mixture Modelling, or any analogous method that
requires ``clusters'' to be ellipses in phase-space (position or
velocity) unsurpsingly yields open clusters that are roughly elliptical.
Some unsupervised clustering methods lead to comparable ``roughly
elliptical'' results \citep[][hereafter
\citetalias{cantatgaudin_gaia_2018}]{cantatgaudin_gaia_2018}.  Other
unsupervised methods such as HDBScan (e.g., \citep[][hereafter
\citetalias{kounkel_untangling_2019}]{kounkel_untangling_2019}) have
been found to yield additional structures, particularly in lower density
regions such as the Psc-Eri stream (CITE: Meingast+18, Curtis+19,
Newton+21), but also more generally around many open clusters
\citep{kounkel_untangling_2019}.  Unsupervised approaches that
incorporate physical constraints like imposing a maximum velocity
dispersion on putative members yield similar results, potentially with
higher purity (Meingast+21).

% TODO: make the bonafide number below quantitative
We've recently been making TESS light curves of age-dated stars across
the sky, as part of a Cluster Difference Imaging Photometry Survey
(CDIPS, CITE Bouma+19).  Our analysis of Cycle 1 (Sectors 1-13) yielded
light curves of \numsouthernuniqlcs candidate cluster members in the
Southern Ecliptic hemisphere, available on
MAST\footnote{\url{https://archive.stsci.edu/hlsp/cdips}}.  Based on
rotation periods, $\approx$25\% appear to be bonafide cluster members.

As part of a broader project of identifying a large and clean sample of
young stars for a transit search, we focus in his paper on a rather
modest question: in just a single rich southern open cluster, is the
cluster halo coeval with the core?  The cluster we chose for this
analysis was NGC\,2516, since it was young ($\sim$120\,Myr) and close
($d\approx400$\,pc) enough to facilitate rotation measurements using
TESS, and some degree of spectroscopic analysis.  We want to know: is
the halo real? To what extent can we use Gaia alone to reliably identify
age-dated needles in the haystack of boring field stars?  And more
generally, what are the implications for the evolution of open clusters
if they do have halos?

Section~\ref{sec:gaia} presents the astrometric data from Gaia, and
clarifies our usage of the terms ``core'' and ``halo''.
Section~\ref{sec:agedate} age-dates the halo of NGC\,2516, based on Gaia
photometry (Section~\ref{subsec:hr}), TESS gyrochronology
(Section~\ref{subsec:tess}), and lithium depletion
(Section~\ref{subsec:lithium}).  In Section~\ref{sec:discussion} we
discuss the implications of this analysis for NGC\,2516 specifically and
stellar spin-down and open cluster evolution generally.
Section~\ref{sec:conclusion} gives our conclusions.

@@

We chose the stars to focus on based on previously reported
memberships of NGC\,2516, available in the literature.  While a number
of pre-Gaia catalogs were available, the purity and accuracy of the
Gaia-derived results are expected to be the current state of the art.
We therefore adopted what we viewed as the most interesting clustering
samples to compare: those of \citet{cantatgaudin_gaia_2018}
(\citetalias{cantatgaudin_gaia_2018}) and
\citet{kounkel_untangling_2019}
(\citetalias{kounkel_untangling_2019})\footnote{A third sample was
also recently reported by \citet{meingast_2021}. Although the results
of this study were made publicly available too late to be included in
our analysis,  1577 of the 1860 \citet{meingast_2021} sources (85\%)
were included in the \citet{kounkel_untangling_2019} sample.  A
complete set of \cn members could very well include some of the newly
reported candidate members from \citet{meingast_2021}.}. While we
could have performed our own clustering analysis based on the Gaia
data, such an effort would be replicating the work of these
investigators. And why would our clustering methods be better than
theirs? We opt instead to use their studies as starting points.

% Theia 613
\citetalias{cantatgaudin_gaia_2018} reported 1106 candidate members of
\cn, brighter than $G=18$\,mag.  \citet{kounkel_untangling_2019}
reported 3003 candidate members, and included stars up to
$\approx$1\,mag fainter.  Each study used unsupervised clustering
based on the second Gaia Data Release, detailed in
Appendix~\ref{app:clustering}.

%TODO calculate these physical sizes carefully
Figure~\ref{fig:gaia6d} shows the cluster members reported by each
study.  The \citetalias{cantatgaudin_gaia_2018} members are all within
a few degrees of the cluster center, while the
\citetalias{kounkel_untangling_2019} members span tens of degrees.  In
physical units, this corresponds to a size difference of XX\,pc versus
YY\,pc.  This difference is somewhat tautological, because
\citetalias{cantatgaudin_gaia_2018} did not extend their search for
members out to tens of degrees from the cluster center, since their
clustering algorithm relied on the contrast between cluster and field
stars in the astrometric space of ($\mu_{\alpha'}, \mu_\delta, \pi$).
Nonetheless, the membership catalog of \citetalias{cantatgaudin_gaia_2018}
echoes that of many previous investigators (CITE, CITE, CITE), and is
consistent with the general {\it visual} impression that one has when
looking at \cn visually: it seems to be at most {\bf X$^\circ$}, and
not a big stream.
This discrepancy begs the question.  What is the true structure of
\cn?  Are the core and halo truly coeval?

To answer these questions, we adopt the
\citetalias{cantatgaudin_gaia_2018} stars as candidate ``core''
members and the \citetalias{kounkel_untangling_2019} stars as
candidate ``halo'' members.  If a star is in both lists, it is a
``core'' member.  1091 of the 1106 core members are in both lists.
This leaves 1912 candidate members in the halo.
Concatenating the two lists yields 3018 candidate members.

In this work, we also define a set of nearby field stars in the
``neighborhood'' of \cn.  Based on the observed distribution of halo
members, we draw these stars randomly from the following ``cube'' in
right ascension, declination, and parallax:
\begin{align}
  $\alpha$\,[{\rm deg}] &\in [108, 132], \\
  $\delta$\,[{\rm deg}] &\in [-76, -45], \\
  $\pi$\,[{\rm mas}] &\in [1.5, 4.0].
\end{align}
We imposed a magnitude limit of $G=19$\,mag, and ran the queries
using the \texttt{astroquery.gaia} module (CITE).
% TODO: why?
We allowed the number of stars in the comparison sample to exceed that
in the cluster sample by a factor of a few, to ensure broad sampling
of stellar masses and evolutionary states.
We also required the comparison sample to not overlap with the cluster
sample, which led to the omission of 1.1\% of the stars drawn over the
volume noted above.

@@

The first check on whether this membership assignment is plausible was
already performed by \citetalias{cantatgaudin_gaia_2018},
\citetalias{kounkel_untangling_2019}, and more recently by
\citet{meingast_2021}.  That check is to see whether the HR diagrams
of the cluster members support the claim that they are coeval.

@@

%
% TODO: the following two paragraphs, when quoted, are ENTIRELY FROM
% the relevant papers. they should probably be nixed in favor of
% concision.
%
\citetalias{cantatgaudin_gaia_2018} applied a procedure that ultimately
yielded what we call ``the core''.  Their procedure was to first query
a Gaia DR2 cone around the previously reported RA and dec of the
cluster, and within $\pm 0.5$\,mas of its previously reported
distance. 
% TODO: which was it?
The outer radius of their cone was either r2 from MWSC (CITE
Kharchenko 2013), or twice the ``cluster radius'' listed by Dias.  No
proper motion cut was applied.  They then applied an unsupervised
classification scheme called UPMASK to $G<18$\,mag stars within this
cone (CITE).  The steps of UPMASK are first to perform a k-means
clustering in the ``astrometric space'' ($\mu_{\alpha'}, \mu_\delta,
\pi$).  Then, a ``veto'' step is applied to assess whether the groups
of stars output from the k-means clustering are or are not more
concentrated than a random distribution.  This is implemented by
``comparing the total branch length of the minimum spanning tree
connecting the stars with the expected branch length for a random
uniform distribution covering the investigated field of view''.  ``To
turn this yes/no flag to a membership probability,
\citet{cantatgaudin_gaia_2018} then redraw new values of
($\mu_{\alpha'}, \mu_\delta, \pi$) for each source based on the listed
value, uncertainty, and covariance.  After a certain number of
redrawings, the final probability is the frequency with which a given
star passes the veto".  In the case of \cn, this yielded a reported
``\texttt{r50}'' within which half of the cluster members were found
to be within 0.496$^\circ$.
When we selected candidate \cn members from the results of
\citetalias{cantatgaudin_gaia_2018}, we opted to include all candidate members
with reported membership probability exceeding 10\%.
While this seems {\it a priori} low, our results ({\bf SECTION XXX})
show that this ``membership probabilty'' severely underestimates the
purity of the \citetalias{cantatgaudin_gaia_2018} sample for \cn. 
%TODO: quantify
Their false positive rate across the sample is more like 1-5\%.

\citetalias{kounkel_untangling_2019}
applied a different unsupervised
clustering method to the 5-dimensional Gaia DR2 data (omitting radial
velocities, due to their sparsity).  Their selection function (see
their Section~2) yielded $\approx 2\times 10^7$ stars, mostly within
$\approx 1$\,kpc and typically with $G<18$\,mag.  Their clustering
algorithm was the ``hierarchical density-based spatial clustering of
applications with noise'' (HDBSCAN, CITE McInnes17).  The classical
DBSCAN algorithm ``identifies clusters as overdensities in a
multi-dimensional space in which the number of sources exceeds the
required minimum number of points within a neighborhood of a
particular linking length $\epsilon$.  HDBSCAN does not depend on
$\epsilon$; instead it condenses the minimum spanning tree by pruning
off the nodes that do not meet the minimum number of sources in a
cluster and reanalyzing the nodes that do. Depending on the chosen
algorithm, it would then either find the most persistent structure
(through the excess of mass method), or return clusters as the leaves
of the tree (which results in somewhat more homogeneous clusters). In
both cases it is more effective at finding structures of varying
densities in a given data set than DBSCAN.''
`` The two main parameters that control HDBSCAN are the number of
sources in a cluster and the number of samples. The former is the
parameter that rejects groupings that are too small; the latter sets
the threshold of how conservative the algorithm is in its
considerations of the background noise (even if the resulting noisy
groupings do meet the minimum cluster size).  By default, the sample
size is set to the same value as the cluster size, but it is possible
to adjust them separately.''
Regarding membership proabilities,
\citetalias{kounkel_untangling_2019} did not report continuous membership
probabilities, instead opting for the binary ``member'' or ``not''.
